# Getting work done {#work}

## Key packages

These packages help me work more efficiently, to the point where I regard them as essential. They'll be used throughout the rest of this document. They don't have any dependencies on other packages, and I expect they'll be available as long as R is. It will be assumed in subsequent sections that these libraries have been loaded:

```{r loadem, echo=FALSE}
library(magrittr)
library(data.table)
```

### Magrittr

Magrittr introduces syntax with "pipes":

```{r magrittr-load, echo=FALSE}
library(magrittr)
```
```{r magrittr-calls, eval=FALSE}
# allowing
x %>% f %>% g %>% h
# instead of 
h(g(f(x)))
```

That is, it allows unnesting of function calls for readability. 

In addition, it allows more compact function definitions (\@ref(function-writing)):

```{r magrittr-funs}
fun = . %>% .^2 %>% sum %>% sqrt
fun(1:3)
```

Despite these advantages, magrittr can greatly slow down code, so avoid using it anywhere where speed might be an issue.

To install, just use CRAN:

```{r inst-mgr, eval=FALSE}
install.packages('magrittr')
```

The name is a pun on [René Magritte's pipe painting](https://en.wikipedia.org/wiki/The_Treachery_of_Images).

### data.table {#data-table}

Data.table offers a variant of the data frame class (from \@ref(data-frames)), optimized for fast sorted and grouped operations and enhanced with cleaner syntax. It also bundles in a variety of other improvements on base R functionality.

#### Modification by reference

In contrast with the rest of R, data.tables are primarily modified "in-place" without assigning the result, which can be much more efficient. As a first example, we can switch the class of a data frame to data.table:

```{r setDT}
DF = data.frame(
  x = letters[c(1, 2, 3, 4, 5)], 
  y = c(1, 2, 3, 4, 5), 
  z = c(1, 2, 3, 4, 5) > 3
)
str(DF)
setDT(DF)
str(DF)
```

We did not use `=` or `<-` to assign the result of `setDT`; the function simply altered `DF` in-place. All of data.table's functions named like `set*` do this.

#### Factor vs character columns

Another difference is that `data.frame` takes string input as a `factor` variable, as seen above; while `data.table` interprets it as character:

```{r DT}
DT = data.table(
  x = letters[c(1, 2, 3, 4, 5)], 
  y = c(1, 2, 3, 4, 5), 
  z = c(1, 2, 3, 4, 5) > 3
)
str(DT)
```

#### Syntax

The third major difference is a change in syntax from a data frame's `DF[i,j]`, with `i` and `j` as indices, to SQL-style syntax that is cleaner for by-group operations:

```{r DTsyntax, eval=FALSE}
# this is only pseudocode:
DT[where, select|update|do, by]
```

We can type column names as barewords here, and even form expressions in terms of columns:

```{r DT-ex}
DT[x > "b", sum(y), by=z]
```

In addition, this call is optimized: The inequality in `i` triggers creation of an "index" in terms of `x`, making subsequent indexing on `x` faster thanks to binary search, documented at `?indices`. And the `sum` in `j` is computed using an optimized "grouped sum" or `gsum` computation, documented at `?GForce`.

The syntax `DT[i,j,by]` should be read as:

1. Subset using `i`
1. Group using `by`
1. Do `j`

The task for `j` can really be anything. For example, going beyond the scope of this document, it's really handy for saving per-group plots:

```{r}
# note: this will save to your current working directory
# type getwd() and read ?setwd for details

bwDT = data.table(MASS::birthwt)

pdf(file="birthweight_graphs.pdf")
bwDT[, {
  case = sprintf("ftv = %s, smoke = %s", .BY$high_vis, .BY$smoke)
  
  cat("Handling", case, "...\n")
  plot(
    age ~ lwt, 
    main = case
  )  
}, by=.(high_vis = ftv >= 1, smoke)]
dev.off()
```

For each group, we're now printing a line with `cat` and saving a plot. This is better than a `for` loop, since we don't have to manually construct some "split-up data" to iterate over; don't have to worry about intermediate variables like `case` contaminating the global environment; and don't have do define what we're doing as a function of prespecified variables -- we can just use any columns of the data.table.

Plotting graphs is beyond the scope of this document, but the example above should make sense after reading the docs for each object, `?MASS::birtwt`, `?pdf`, et al.

#### Installation

Again, CRAN can be used:

```{r inst-dt, eval=FALSE}
install.packages('data.table')
```

However, the package is under active development, with new features available only in the development version. To install it, follow [the instructions from the package wiki](https://github.com/Rdatatable/data.table/wiki/Installation):

```{r inst-dt-dev, eval=FALSE}
remove.packages("data.table")
install.packages("data.table", type = "source",
  repos = "http://Rdatatable.github.io/data.table")
```

#### Getting started

[The official vignettes](https://github.com/Rdatatable/data.table/wiki/Getting-started) for the package are a great way to start with the package. I started by reading some slides and the FAQ in full before using the package. 

I also found it helpful to turn on verbose output when first learning:

```{r verbose, eval=FALSE}
options(datatable.verbose=TRUE)
```

This makes a variety of data.table operations more transparent by means of more detailed messages in the console. See `?data.table` and `?options` for details.

## Building sequences

Use a single colon to build a sequence of integers:

```{r seq-colon}
3:5
```

### Counting upwards

If we are running `1..n`, it is safer and more efficient to use `seq_len` than `1:n`:

```{r seq-len, eval=FALSE}
n_x = 1e5
n_y = 1e5
n   = x + y
z   = seq_len(n)
```

It is safer in the sense that, were `n` zero, we would get the correct result of a zero-length vector, while `1:0` gives `c(1L, 0L)`.

If we are running a sequence `1...` alongside a vector, then `seq_along` is the right tool:

```{r seq-along}
x = c(3, 3, 4)
seq_along(x)
```

<span class="spdetails"> **R function names.**
It is not worth trying to make sense of R's naming conventions. Presumably thanks to historical accidents (some involving R's ancestor, S), we have names like `seq.int`, `seq_along`, `setNames` and `Recall`.
</span>

### Fancier sequences

From here, the variety of options start to resemble what we saw for the `rep` function in \@ref(initializing).

The function `seq.int` extends the colon operator:

```{r seq-int}
seq.int(5, 10)
seq.int(5, 10, by = 2)
seq.int(5, by = 3, length.out = 3)
seq.int(to = 100, by = -11, length.out = 3)
```

The `seq` function extends further by allowing for non-integer values:

```{r seq}
seq(5, 10, by=.5)
```

## Subsetting {#dt-subset}

### Selecting by value

This is analogous to a SQL WHERE clause or a Stata `if` clause.

```{r dt-sub-pre}
# example data
DT = data.table(Titanic)
```

```{r dt-sub}
DT[ Class == "Crew" & N > 100 ]
```

This case, `DT[i]`, is the subsetting task needed in the vast majority of cases. The rest of \@ref(dt-subset) can safely be skipped the first time through; it is mostly useful for reference.

### Selecting by group statistic

This is analogous to a SQL HAVING clause.

```{r dt-sub-gstat}
DT[ , if (sum(N) < 300) .SD, by=Class ]
```

We are including the group only if it meets our condition. The structure `if (cond) x` returns `x` if the condition is true and nothing (`NULL`) otherwise, omitting those rows. `.SD` refers to the Subset of the Data associated with each `by=` group.

This syntax isn't as clean as a HAVING clause, but it is likely to improve.

### Selecting rows within each group

```{r dt-sub-grow}
# select first row
DT[ , .SD[1L], by=Class ]
# select row(s) with highest count
DT[ , .SD[N == max(N)], by=Class ]
```

<span class="spdetails"> **Efficient selection of rows within each group.** 
The second task here is [somewhat inefficient](https://github.com/Rdatatable/data.table/issues/735) currently, with a faster workaround [provided by eddi](http://stackoverflow.com/a/16574176): `DT[DT[ , .I[N == max(N)], by=Class ]$V1]`. The variable `.I` represents the row number in the full table, or at least it *should* do that. Currently, `.I` does [not always behave like this](https://github.com/Rdatatable/data.table/issues/1494).
</span>

## Summary statistics {#dt-summary-stats}

### Counting rows {#counting-rows}

bwDT[, .N, by=ftv]

refer to \@ref(tallying)

### The `summary` function



### Create aggregated table

collapse


### Save to current table





## Modifying data in-place {#dt-subassign}

Creating, editing and removing columns are all done using `:=` in `j` of `DT[i, j, by]`. This functionality operates in-place, in the sense that the underlying data stays in the same place.

<span class="spdetails">
**What can be modified in-place?** The scope of in-place operations is currently limited to altering columns. Adding and removing rows by reference is not yet supported, since this is harder to do in R, thanks to its column-oriented storage of tables (contrasting with many database systems that store data rowwise for easy INSERT and DELETE queries).
</span>

### Creating columns {#dt-col-create}

```{r dt-add}
DT = data.table(
  x = letters[c(1, 2, 3, 4, 5)], 
  y = c(1, 2, 3, 4, 5), 
  z = c(1, 2, 3, 4, 5) > 3
)
# creating a column
DT[, u := 5:1][]
# creating multiple
DT[, `:=`(v = 2, w = 3L)][]
# creating with dynamic names
nms = c("a", "b", "c")
DT[, (nms) := .(1, 2, 3)][]
```

All of these tasks are performed in-place -- altering `DT` without making a new object. Usually, the results are not displayed. They appear here because `[]` is "chained" onto the end of each task.

<span class="spnote">
**`:=` is the function both for creation and modification of columns.** This will be covered it more detail in subsequent sections, but is worth emphasising. In particular, this contrasts with Stata.
</span>

### Removing columns

```{r dt-remove}
nms = c("u", "v", "w", "a", "b", "c")
DT[, (nms) := NULL][]
```

A warning will print if you remove any columns that don't currently exist.

### Replacing entire columns

Data.table is careful about column types:

```{r dt-reptot, warning=TRUE}
DT[, a := 10L ][]
DT[, a := 21L ][]
DT[, a := 32.5 ][]
```

It knows that `a` is of integer type and sees that 32.5 is conspiciously not an integer and so gives a warning when coercing 32.5 to an integer (to match `a`). Elsewhere in R, in contrast, `a` would be coerced to match the float `32.5` -- probably not the behavior we want -- with no warning.

As indicated in the warning message, if we want to change `a`'s type, we have to explicitly replace all values:

```{r dt-reptot-class, warning=TRUE}
DT[, a := rep(32.5, .N) ][]
```

The coercion is done silently, but it can be made more visible by turning on `verbose`, which will refer to the "plonking" of a new vector in place of the old.

### Replacing columns conditionally {#dt-ifelse}

This is analogous to a SQL UPDATE query or a replace statement in Stata. To illustrate, we will look again at a vectorized `if`/`else` assignment, mentioned in \@ref(ifelse):

```{r dt-ifelse}
DT[     , b := "Ants"] # initialize to baseline value
DT[y > 1, b := "Bats"] # replace based on a condition
```

This can also be done with chaining:

```{r dt-ifelse-chain}
DT[, b := "Ants"][y > 1, b := "Bats"]
```

This chaining works because `DT[...][...]` is evaluated like `(DT[...])[...]` and the return value of the piece in parentheses is a data.table.

As we saw in the last section, partial replacement of a column will trigger a warning if the classes don't match:

```{r dt-ifelse-conflict, warning=TRUE}
DT[, b := "Ants"][y > 1, b := 111]
```

Another nice feature, similar to Stata, is reporting of the number of rows modified. This can be seen by turning `verbose` on. (Unfortunately, as of August 2016, this display has [a small departure from Stata's behavior](https://github.com/Rdatatable/data.table/issues/1808).)

### Iterating over columns

Suppose we want to find the maximum value in each column. As we saw in \@ref(lapply-df), we can do this like `lapply(DT, max)`. The `lapply` function is an option here because a data.table, like a data frame, is a list of columns.

With data.table, however, we can go beyond a simple `lapply`. We can, for example, compute the max within each value of `z`, limiting computation to numeric columns:

```{r lapply-dt}
DT[, 
   lapply(.SD, max)
, by=z, .SDcols = { sapply(DT, is.numeric) }]
```

`.SDcols` is used to filter the columns that appear in `.SD`, the subset of data available in `j`. 

### Another interface, `set`

The `set` function can also be used to modify data by reference. It has efficiency benefits in some cases, but it essentially a more limited version of `:=`.

### Avoiding in-place modification

If you want to create a new data.table instead of modifying in-place, use

```{r dt-copy, eval=FALSE}
DT2 = copy(DT1)
```

We have to use this instead of `DT2 = DT1` since with the latter we have only created a new "pointer" to the first data set. I routinely make copies after reading in data so that I can back out where I tripped over something in the process of data cleaning.

## Joins {#dt-merging}

merge-assign + transform / gen http://stackoverflow.com/q/37067905/

more merge-assign http://stackoverflow.com/a/37440011/


warn don't switch(), merge, consider using stack, note that match() is basically a merge http://stackoverflow.com/a/36755530/

"joins" or "merges"


setkey

X[Y, j] + i.*

ex: ...

X[Y, j, by=.EACHI] + on

X[!Y] + which

rbind, rbindlist



## Managing data types

use integers when you can, use characters or integers for keys

use characters when it makes sense, don't worry about memory. the real memory problems come from how data is organized, as discussed in \@ref(structuring-data).

remember: do rounding as late as possible -- don't start during the cleaning and exploration phase


### Dates and times {#dates-times}

### List columns

list columns -- somewhat clumsy, but sometimes convenient (no by=, shift(); inefficient %in% and match()). can store regression results, for example

consider unique %>% sort %>% toString/paste instead



## Managing multiple data sets {#structuring-data}

11 structuring data: keys, merging and reshaping (db_design)

background: tidy data / database design

unique, duplicated + fromLast, CJ + unique




setdiff, union, intersect (works with data.table from 1.9.8+)

### Browsing loaded objects {#tables}

ls()

rm()

tables()


## Exploring data {#exploring-data}

range, summary

Henk Harmsen on "ergonomics" https://rpubs.com/carbonmetrics/datatable_ihub

`e_bad = DT[, var, by=e_id][var==TRUE, e_id]; DT[.(e_id)] (simpler with having)`

`DT[, .N, by=e_id][order(-N)]`

creating temp vars, printing with []

DT[, do_stuff, j] %>% print(nrow=Inf) with magrittr (warning about performance hit in real nonexploratory code)

hist, hexbin

cor, cormat

exploring models: curve

qqplot etc for the statistically literate, and other tools for ML people

### Reshaping


## Input and output {#input-output}

warn forward slashes in file names


fread

tstrsplit

as.IDate, IDateTime

wday, weekdays, etc.

show that max, min, inequalities work for dates and times

show how ordered factors can be used to set up inequalities for strings

lapply + .SDcols with date conversion

shortcuts: .SDcols = V1:V10, .SDcols = grep("^date_", names(DT)), etc

state.abb, state.names

write.csv

readxl

writeXLS

use google, these I/O functions will always be in active development

haven is a thing for Stata files, but fuggedaboutit unless you can't access Stata. far better to use csvs for data transfer, metadata be damned

with Rdata, use load() or just drag it into the window




## Approximation, optimization and regression

(so far, we've just talked about data manipulation, now onto math/stats)

obvs, if your problem has a closed form solution, do the matrix algebra and don't get bogged down in syntax oddities of a particular language

seriously consider using a matrix or array if it fits the problem

integration

optimization

formula interface

dynamically constructing formulae -- too much trouble

precomputing indices http://stackoverflow.com/q/19279075/1191259
