# Getting work done {#work}

## Key packages

These packages help with working more efficiently, to the point where I regard them as essential. They'll be used throughout the rest of this document.

They don't have any dependencies on other packages, and I expect they'll be available as long as R is. It will be assumed in subsequent sections that these libraries have been loaded:

```{r loadem, echo=FALSE}
library(magrittr)
library(data.table)
```

### Magrittr

Magrittr introduces syntax with "pipes":

```{r magrittr-load, echo=FALSE}
library(magrittr)
```
```{r magrittr-calls, eval=FALSE}
# allowing
x %>% f %>% g %>% h
# instead of 
h(g(f(x)))
```

That is, it allows unnesting of function calls for readability. 

In addition, it allows more compact function definitions (\@ref(function-writing)):

```{r magrittr-funs}
fun = . %>% .^2 %>% sum %>% sqrt
fun(1:3)
```

Despite these advantages, magrittr can greatly slow down code, so avoid using it anywhere where speed might be an issue.

To install, just use CRAN:

```{r inst-mgr, eval=FALSE}
install.packages('magrittr')
```

The name is a pun on [René Magritte's pipe painting](https://en.wikipedia.org/wiki/The_Treachery_of_Images).

### data.table {#data-table}

Data.table offers a variant of the data frame class (from \@ref(data-frames)), optimized for fast sorted and grouped operations and enhanced with cleaner syntax. It also bundles in a variety of other improvements on base R functionality.

#### Modification by reference

In contrast with the rest of R, data.tables are primarily modified "in-place" without assigning the result, which can be much more efficient. As a first example, we can switch the class of a data frame to data.table:

```{r setDT}
DF = data.frame(
  x = letters[c(1, 2, 3, 4, 5)], 
  y = c(1, 2, 3, 4, 5), 
  z = c(1, 2, 3, 4, 5) > 3
)
str(DF)
setDT(DF)
str(DF)
```

We did not use `=` or `<-` to assign the result of `setDT`; the function simply altered `DF` in-place. All of data.table's functions named like `set*` do this.

#### Factor vs character columns

Another difference is that `data.frame` takes string input as a `factor` variable, as seen above; while `data.table` interprets it as character:

```{r DT}
DT = data.table(
  x = letters[c(1, 2, 3, 4, 5)], 
  y = c(1, 2, 3, 4, 5), 
  z = c(1, 2, 3, 4, 5) > 3
)
str(DT)
```

#### Syntax

The third major difference is a change in syntax from `DF[i,j]` with `i` and `j` as indices to SQL-style syntax that is cleaner for by-group operations:

```{r DTsyntax, eval=FALSE}
# pseudocode
DT[where, select|update|do, by]
```

We can type column names as barewords here, and even form expressions in terms of columns:

```{r DT-ex}
DT[x > "b", sum(y), by=z]
```

In addition, this call is optimized:

- The inequality in `i` triggers creation of an "index" in terms of `x`, making subsequent indexing on `x` faster thanks to binary search, documented at `?indices`.
- The `sum` in `j` is computed using an optimized "grouped sum" or `gsum` computation, documented at `?GForce`.

#### Generality

so much better than by: do

x := {make_x}

{} for general tasks

custom by, NSE in all args

#### Installation

Again, CRAN can be used:

```{r inst-dt, eval=FALSE}
install.packages('data.table')
```

However, the package is under active development, with new features available only in the development version. To install it, follow [the instructions from the package wiki](https://github.com/Rdatatable/data.table/wiki/Installation):

```{r inst-dt-dev, eval=FALSE}
remove.packages("data.table")
install.packages("data.table", type = "source",
  repos = "http://Rdatatable.github.io/data.table")
```

#### Getting started

[The official vignettes](https://github.com/Rdatatable/data.table/wiki/Getting-started) for the package are a great way to start with the package. I started by reading some slides and the FAQ in full before using the package. 

I also found it helpful to turn on verbose output when first learning:

```{r verbose, eval=FALSE}
options(datatable.verbose=TRUE)
```

This makes a variety of data.table operations more transparent by means of more detailed messages in the console. See `?data.table` and `?options` for details.

## Building sequences

Use a single colon to build a sequence of integers:

```{r seq-colon}
3:5
```

### Counting upwards

If we are running `1..n`, it is safer and more efficient to use `seq_len` than `1:n`:

```{r seq-len, eval=FALSE}
n_x = 1e5
n_y = 1e5
n   = x + y
z   = seq_len(n)
```

It is safer in the sense that, were `n` zero, we would get the correct result of a zero-length vector, while `1:0` gives `c(1L, 0L)`.

If we are running a sequence `1...` alongside a vector, then `seq_along` is the right tool:

```{r seq-along}
x = c(3, 3, 4)
seq_along(x)
```

<span class="spdetails"> **R function names.**
It is not worth trying to make sense of R's naming conventions. Presumably thanks to historical accidents (some involving R's ancestor, S), we have names like `seq.int`, `seq_along`, `setNames` and `Recall`.
</span>

### Fancier sequences

From here, the variety of options start to resemble what we saw for the `rep` function in \@ref(initializing).

The function `seq.int` extends the colon operator:

```{r seq-int}
seq.int(5, 10)
seq.int(5, 10, by = 2)
seq.int(5, by = 3, length.out = 3)
seq.int(to = 100, by = -11, length.out = 3)
```

The `seq` function extends further by allowing for non-integer values:

```{r seq}
seq(5, 10, by=.5)
```

## Subsetting data {#dt-subset}

The syntax `DT[i,j,by]` can be read as:

1. Subset using `i`
1. Group using `by`
1. Do `j`

### Selecting by value

This is analogous to a SQL WHERE clause.

```{r dt-sub-pre}
# example data
DT = data.table(Titanic)
```

```{r dt-sub}
DT[ Class == "Crew" & N > 100 ]
```

This is the subsetting task needed in the vast majority of cases. The reader can probably skip the rest of \@ref(dt-subset) the first time through; it's mostly useful for reference.

### Selecting by group statistic

This is analogous to a SQL HAVING clause.

```{r dt-sub-gstat}
DT[ , if (sum(N) < 300) .SD, by=Class ]
```

`.SD` refers to the Subset of the Data associated with each `by=` group. The structure `if (cond) x` returns `x` if the condition is true and nothing (`NULL`) otherwise, omitting those rows.

This isn't as clean as a HAVING clause, but is the current idiom for it.

### Selecting rows within each group

```{r dt-sub-grow}
# select first row
DT[ , .SD[1L], by=Class ]
# select row(s) with highest count
DT[ , .SD[N == max(N)], by=Class ]
```

The second task here is [somewhat inefficient](https://github.com/Rdatatable/data.table/issues/735) currently, with a faster workaround [provided by @eddi](http://stackoverflow.com/a/16574176):

```{r dt-sub-grow-eff}
# select row(s) with highest count
DT[DT[ , .I[N == max(N)], by=Class ]$V1]
```

`.I` is the row number in the full table. Run the code in pieces to see how it works:

- `DT[ , .I[N == max(N)], by=Class ]` contains the filtered row numbers by group
- `DT[ , .I[N == max(N)], by=Class ]$V1` is a row index vector

<span class="spnote">
**Quirk of `.I`.** The variable `.I` *should* always refer to rows of the full table, but currently only does this is `by=` is given. Compare `DT[ Survived == "Yes", .I]` to `DT[ Survived == "Yes", .I, by=Sex]$I`.
</span>

## Modifying data {#dt-cols}

Creating, editing and removing columns are all done with `:=` in `j` of `DT[i, j, by]`.

### Creating columns

```{r dt-add}
DT = data.table(
  x = letters[c(1, 2, 3, 4, 5)], 
  y = c(1, 2, 3, 4, 5), 
  z = c(1, 2, 3, 4, 5) > 3
)
# creating a column
DT[, u := 5:1][]
# creating multiple
DT[, `:=`(v = 2, w = 3L)][]
# creating with dynamic names
nms = c("a", "b", "c")
DT[, (nms) := .(1, 2, 3)][]
```

All of these tasks are performed in-place and so do not print their results by default. Here, the result is being displayed because `[]` is "chained" onto the end of each operation.

### Removing columns

```{r dt-remove}
nms = c("u", "v", "w", "a", "b", "c")
DT[, (nms) := NULL][]
```

If asked to remove columns that don't exist, the call will print a warning.

### Replacing entire columns

Data.table is careful about column types:

```{r dt-reptot, warning=TRUE}
DT[, a := 10L ][]
DT[, a := 21L ][]
DT[, a := 32.5 ][]
```

It knows that `a` is of integer type and sees that 32.5 is conspiciously not an integer and so gives a warning when coercing 32.5 to an integer (to match `a`). Elsewhere in R, in contrast, `a` would be coerced to match the float `32.5` -- with no warning.

As indicated in the warning message, if we want to change `a`'s type, we have to explicitly replace all values:

```{r dt-reptot-class, warning=TRUE}
DT[, a := rep(32.5, .N) ][]
```

Here, the coercion is done silently, but it can be made more visible by turning on `verbose`, which will refer to the "plonking" of a new vector in place of the old.

### Replacing columns conditionally {#dt-ifelse}

This is analogous to a SQL UPDATE clause or a replace statement in Stata.

warns about coercion

verbose indicates number of rows modified, protection against invalid indexing

set alternate interface for modifying columns

"replace", "update"

advantages over vanilla subassign from \@ref(subassigning): preserves class unless totally overwriting, can report number of changes with verbose to avoid invalid indexing

better ifelse

merge-assign + transform / gen http://stackoverflow.com/q/37067905/

more merge-assign http://stackoverflow.com/a/37440011/

### Iterating over columns

.SDcols = sapply(DT,class)=="character"


## Operations by group {#dt-grouping}

### Splitting data



### Summary statistics {#dt-summary-stats}

save to current table

aggregation

tallying, point to next chapter


## Joins {#dt-merging}


warn don't switch(), merge, consider using stack, note that match() is basically a merge http://stackoverflow.com/a/36755530/

"joins" or "merges"

use copy(), i do for a backup while working

## Modification in-place

ex: modification by reference even happens from within a function

## Inspecting all objects

tables() or jan's extension; str(); sapply(DT, class)



## Structuring data {#structuring-data}
### Dates and times {#dates-times}

11 structuring data: keys, merging and reshaping (db_design)

background: tidy data / database design

unique, duplicated + fromLast, CJ + unique

use integers when you can, use characters or integers for keys

remember: do rounding as late as possible -- don't start during the cleaning and exploration phase

setkey

X[Y, j] + i.*

ex: ...

X[Y, j, by=.EACHI] + on

X[!Y] + which

rbind, rbindlist

setdiff, union, intersect (works with data.table from 1.9.8+)

sort vs order

cumsum, diff, rev, rleid

list columns -- somewhat clumsy, but sometimes convenient (no by=, shift(); inefficient %in% and match()). can store regression results, for example

consider unique %>% sort %>% toString/paste instead

reference: data.table vignettes

## Exploring data {#exploring-data}

range, summary

Henk Harmsen on "ergonomics" https://rpubs.com/carbonmetrics/datatable_ihub

`e_bad = DT[, var, by=e_id][var==TRUE, e_id]; DT[.(e_id)] (simpler with having)`

`DT[, .N, by=e_id][order(-N)]`

creating temp vars, printing with []

DT[, do_stuff, j] %>% print(nrow=Inf) with magrittr (warning about performance hit in real nonexploratory code)

hist, hexbin

cor, cormat

exploring models: curve

qqplot etc for the statistically literate, and other tools for ML people

## Input and output {#input-output}

warn forward slashes in file names


fread

tstrsplit

as.IDate, IDateTime

wday, weekdays, etc.

show that max, min, inequalities work for dates and times

show how ordered factors can be used to set up inequalities for strings

lapply + .SDcols with date conversion

shortcuts: .SDcols = V1:V10, .SDcols = grep("^date_", names(DT)), etc

state.abb, state.names

write.csv

readxl

writeXLS

use google, these I/O functions will always be in active development

haven is a thing for Stata files, but fuggedaboutit unless you can't access Stata. far better to use csvs for data transfer, metadata be damned

with Rdata, use load() or just drag it into the window

## Browsing loaded objects {#tables}

ls()

rm()

tables()



## Approximation, optimization and regression

(so far, we've just talked about data manipulation, now onto math/stats)

obvs, if your problem has a closed form solution, do the matrix algebra and don't get bogged down in syntax oddities of a particular language

seriously consider using a matrix or array if it fits the problem

integration

optimization

formula interface

dynamically constructing formulae -- too much trouble

precomputing indices http://stackoverflow.com/q/19279075/1191259
